Lower Bounds on the Minimax Risk for the \\ Source Localization Problem

### 2017-01-25 paper.tex
Make major edits to foreshadow sensor size issue.

@@

The one-dimensional toy problem is an effective tool for understanding bounds
on source localization accuracy in other settings as well. A vast literature on
linear inverse problems and deconvolution algorithms exists
(see~\cite{Bal2012Introduction} for an introduction to this field), and has
addressed the reduced one-dimensional problem in broad settings
(see~\cite{Cavalier2002Sharp,Efromovich1997Robust,Ibragimov1981Statistical} and
references therein). However, our setting and interpretation appears to be
unique, since most prior work focuses on recovering a whole function (given
certain smoothness constraints), rather than locating a point
source~\cite{Cavalier2002Sharp}. Work that \emph{does} address point sources
and location-based error metrics~\cite{Ibragimov1981Statistical} does not, to
our knowledge, address scaling in the number of sensors.

Hence, we believe that this paper is the first to give lower bounds on source
localization error (measured in an $\ell_2$-distance sense) in estimating the
location of a one-dimensional point source, which is observed by sensors
through a diffusive medium (treated as a low-pass filter), and corrupted by
additive white Gaussian noise. We describe the problem setting in detail in
Section~\ref{sec:source-localization} and the minimax techniques and our main
results (for a naive and then for a more physical sensor model) in
Section~\ref{sec:minimax-lower-bounds}. We conclude in
Section~\ref{sec:discussion} with a discussion on the goodness of our bounds,
and on how our results can be extended to other simple settings.

@@

The sensors sample the shifted continuous-space impulse response, with some
additive noise. We denote the noiseless sampled version of $x(s; \theta)$ by
the $m$-length vector $\v x(\theta)$:
\begin{equation} \label{eq:sampled-signal}
	\v x(\theta) = \bigg[x(0; \theta), \ldots, x\Big(\frac{kS}{m}; \theta\Big), \ldots, x\Big(\frac{(m-1)S}{m}; \theta\Big)\bigg]^T
\end{equation}
where $m$ is the number of sensors and $k \in \{0, 1, \ldots, m-1\}$. The
additive noise is given by $\v \epsilon$ (to be described shortly), and the
noisy samples are represented by
\begin{equation} \label{eq:sensor-obs}
	\v y = \v x(\theta) + \v \epsilon.
\end{equation}
In the source localization problem, we receive $n$ trials\footnote{The word
``trials'' comes from neuroscience literature, where several trials of an
experiment are run with a single participant. The participant is given the same
stimulus in each trial, evoking the same response in the brain (usually with
some trial-to-trial variability). For our purposes, we assume that trials are
effectively i.i.d.\ realizations of sensor noise, with no trial-to-trial
variability.} (i.e., $n$ realizations) of $\v y$ according to
equation~\eqref{eq:sensor-obs}, for the same point source $\theta$, using which
we need to estimate the location of the source.
The complete signal model described in this section is summarized in
Fig.~\ref{fig:signal-model}.

\subsection{Channel model}
\label{sec:channel-model}

The noise $\v \epsilon$, introduced in equation~\eqref{eq:sensor-obs}, is the
only source of uncertainty in the problem. For our purposes, we assume that $\v
\epsilon$ is zero-mean and Gaussian, $\v \epsilon \sim \mathcal{N}(\v 0, \m
\Sigma)$. We restrict our analysis to a ``sensor noise'' setting, wherein
$\m\Sigma = \sigma^2 \m I$, i.e., noise is i.i.d.\ across sensors.

The noise is also assumed to be i.i.d.\ across trials, i.e., the $n$
realizations of $\v y$ are independently drawn from $P(\theta)$, for the
same location parameter $\theta$.

@@

Thus, the final lower bound can be written as:
\begin{equation} \label{eq:le-cam-bound}
	\mathfrak{M}_n(\mathcal P, \Phiorho) \geq \frac{\Phi(\delta)}{2} \big(1 - \norm{P_1^n - P_0^n}_{TV}\big)
\end{equation}
The superscripts ``$n$'' remind us that these are $n$-fold product
distributions, since we have $n$ i.i.d.\ trials used in the estimate.  The
difficulty in Le Cam's method lies in selecting the two hypotheses to trade-off
the effect due to a small value of $\delta$ and a large value of $\norm{P_1^n -
P_0^n}_{TV}$ appropriately, to derive the tightest possible bound.

@@

For the source localization problem, we measure \emph{risk} in terms of the
squared-distance between the true location and the estimated location. Hence,
our loss function is $\Phi(\rho(\theta, \hat\theta)) = \abs{\theta -
\hat\theta}^2$, where $\hat\theta$ is some estimate of the location based on
$n$ trials of the noisy sensor observations $\v y$.

@@

We now state the main result of this paper.

@@

@@

@@

section~\ref{sec:channel-model}. Hence, combining

@@

	\norm{\v\Delta}^2 \leq 4 \kappa^2 \delta^2 \floor*{\frac{(w + 2\delta)m}{S} + 1} \leq 4 \kappa^2 \delta^2 \Big(\frac{(w + 2\delta) m}{S} + 1\Big).

@@

\subsection{Towards a more physical sensor model}

The bound in Theorem~\ref{thm:main-lb} may mislead one to think that the error
in source localization goes to zero, as the number of sensors goes to infinity.
This observation is contradicts the information-theoretic lower bound given
in~\cite{Grover2016Fundamental}, where even with an infinite number of sensors,
the lower bound is strictly greater than zero. This indicates that the bound in
section~\ref{sec:lecam-lb} is failing to capture some effect which leads to
saturation of error as the number of sensors increases.

Indeed, the bound in Theorem~\ref{thm:main-lb} assumes that an arbitrarily
large number of sensors can be squeezed into a finite amount of space, and that
these sensors \emph{still} have the same SNR.  This is non-physical, since in
reality, each sensor has a certain width, and the value measured by it is the
integral of the continuous-space signal within that width (analogous to a
sample-and-hold circuit, which integrates a continuous-\emph{time} signal over
a short period).  If the width of a sensor is decreased, then effectively, its
SNR reduces.  This is intuitively evident in the sample-and-hold analogy: to
compute the signal value, the integral must be divided by the time interval
over which the integration is performed; this division amplifies noise as the
time interval gets smaller.

We performed a preliminary study to understand the effect of viewing sensors as
integrators. In the new sensor model, sensors have a width which decreases as
their number grows to infinity. Concretely, the $k$-th sensor which is located
at $s = kS/m$ extends from $(k{-}\frac{1}{2})S/m$ to $(k{+}\frac{1}{2})S/m$.
The signal sensed by this sensor is the $k$-th element of $\v x(\theta)$, now
given by
\begin{equation}
	x_k(\theta) = \int_{(k-\frac{1}{2})S/m}^{(k+\frac{1}{2})S/m} x(s;\theta) ds
\end{equation}
The noise vector $\v \epsilon$ is also generated similarly, by integrating over
small sections of a white noise process, which is assumed to additively corrupt
the continuous-space signal $x(s;\theta)$ prior to sampling. This leads to
i.i.d.\ noise values at sensors (since they do not overlap), with variance
inversely proportional to $m$, i.e., $\sigma(m) \propto 1/\sqrt{m}$.

The bounding technique used in section~\ref{sec:lecam-lb} no longer easily
admits a closed-form solution which gives a bound for all values of $m$. We
instead compute the bound numerically, using exact expressions for error
probability in a binary hypothesis test. For Gaussian distributions of equal
variance and a uniform prior on the hypotheses, the minimum possible error
probability (over all tests) is a $Q$-function in the distance between their
means~\cite{NPLemmaOrSimilar},
\begin{equation}
	P_e = \inf_\psi \mathbb P(\psi(Y^n) \neq V) = Q\bigg(\frac{\norm{\v\mu_0 - \v\mu_1}}{2\sigma(m)}\bigg).
\end{equation}
where $Q(x) = \int_x^\infty \frac{1}{\sqrt{2\pi}} e^{-u^2/2} du$. Hence, from
Theorem~\ref{thm:est-to-testing},
\begin{equation}
	\mathfrak{M}_n(\mathcal{P}, \Phiorho) \geq \Phi(\delta) P_e = \delta^2 Q\bigg(\frac{\norm{\v x(\theta_0) - \v x(\theta_1)}}{2\sigma(m)}\bigg).
\end{equation}
This bound is valid for any $\{\theta_0, \theta_1\}$ constituting a
$2\delta$-packing of $\Theta$. To numerically optimize over all possible
packings, we relax the process by fixing $\theta_0 = 0$ and optimizing over
$\theta_1 = 2\delta$.  We compute sensor values by performing numerical
integration, and find the best possible $\delta$ by grid search.

\begin{figure}[t]
	\centering
	\includegraphics[width=2.8in]{lb-vs-m-annotated}
	\caption{A numerically generated plot of the lower bound on minimax risk of
		source localization, against the number of uniformly distributed
		sensors, $m$, used to estimate the location. The different curves are
		for different numbers of ``trials'' $n$ (i.i.d.\ sensor observations).
		The bound saturates for large $m$, indicating that for a fixed number
		of trials (or equivalently, for a fixed SNR), there is reduced benefit
		in increasing the number of sensors beyond a point. However, with more
		trials, the bound saturates at increasingly larger numbers of sensors.
	Thus, there is benefit to increasing the number of sensors, provided SNR
can be increased by leveraging larger numbers of trials.}
	\label{fig:numerical}
\end{figure}

Plotting this numerically computed bound for different numbers of trials (see
Fig.~\ref{fig:numerical}) reveals some interesting trends. First, the bound
saturates as $m$ grows large, matching the behaviour presented
in~\cite{Grover2016Fundamental}. Second, the lower bound on error is larger for
smaller numbers of sensors, as we expect. Third and most importantly, as the
number of trials increases, bound begins to saturate only for larger numbers of
sensors. In other words, there is increased benefit in using larger numbers of
sensors, when we also have more trials\footnote{Recall that the number of
``trials'' is the number of i.i.d.\ sensor observations we receive, per
equation~\ref{eq:sensor-obs}.}. This substantiates the intuition given
in~\cite{Grover2016Information}, which addresses this issue using a
Nyquist-rate perspective. A larger number of sensors will sample a larger
number of high-frequency components of the source signal (and hence allow us to
locate it better), but only if the noise floor is sufficiently low. Noise
variance, in turn, can be reduced by averaging over multiple trials.

@@

While the bounds in section~\ref{sec:lecam-lb} were derived for a
one-dimensional circular domain, extensions to a one-dimensional linear domain
(of fixed length) and to a $d$-dimensional domain (circular or linear) are
fairly straightforward, under certain conditions. For a 1D linear domain, we
require that sensors capture the entirety of the impulse response even at the
edges of the domain. Hence, sensors must be placed uniformly in $[-w/2,
S{+}w/2]$. For extending to $d$ dimensions, it suffices to apply Assouad's
Lemma~\cite{Tsybakov2009Introduction} to separate the problem into $d$
independent one-dimensional problems. Such extensions substantially broaden the
number of settings in which this bound applies.

Another point on sensor placement bears repeating: the bounds derived in this
paper apply to only a specific sensor configuration, i.e., the uniform one.
Intuitively, this is the correct choice in the minimax setting, since any
non-uniform placement will have one or more points which is far from all
sensors. The ``maximization'' over $\theta \in \Theta$ would then select a
$\theta$ for which the value of the impulse response at the sensors is
smallest, so that the ratio of mean-sensor-value to noise variance (a form of
SNR) is minimized.

@@

>>>>

\setlength{\textfloatsep}{10pt}

@@

The one-dimensional toy problem is an effective tool for understanding bounds
on source localization accuracy in other settings as well. A vast literature on
linear inverse problems and deconvolution algorithms exists
(see~\cite{Bal2012Introduction} for an introduction to this field), and has
addressed the reduced one-dimensional problem in broad settings
(see~\cite{Cavalier2002Sharp,Efromovich1997Robust,Ibragimov1981Statistical} and
references therein). However, our setting and interpretation appears to be
unique, since most prior work focuses on recovering a whole function (given
certain smoothness constraints), rather than locating a point
source~\cite{Cavalier2002Sharp}. Work that \emph{does} address point sources
and location-based error metrics~\cite[Ch.~7, Sec.~2]{Ibragimov1981Statistical}
does not, to our knowledge, address scaling in the number of sensors.

Hence, we believe that this paper is the first to give lower bounds on source
localization error (measured in an $\ell_2$-distance sense) in estimating the
location of a one-dimensional point source, which is observed by sensors
through a diffusive medium (treated as a low-pass filter), and corrupted by
additive white Gaussian noise. We are also the first to analyze the
implications physical sensor models on such lower bounds. A simplistic
theoretical model might assume that the number of sensors can increase
unbounded, while maintaining a fixed SNR at each sensor. We show that this is
not physically possible in our setting, and derive numerical bounds which have
radically different asymptotic behaviour.

We describe the problem setting in detail in
Section~\ref{sec:source-localization} and the minimax techniques and our main
results (for a na\"ive and then for a more physical sensor model) in
Section~\ref{sec:minimax-lower-bounds}. We conclude in
Section~\ref{sec:discussion} with a discussion on the goodness of our bounds,
and on how our results can be extended to other simple settings.

@@

\subsection{Sensor and noise models}
\label{sec:sensor-model}

We consider two sensor models: a simplistic ``point sensor'' model and an
``integrator sensor'' model, to be used in sections~\ref{sec:lecam-lb}
and~\ref{sec:physical-model} respectively.

In the ``point sensor model'', sensors \emph{sample} the shifted
continuous-space impulse response, with some additive noise. We denote the
noiseless sampled version of $x(s; \theta)$ by the $m$-length vector $\v
x(\theta)$,
\begin{equation} \label{eq:sampled-signal}
	\v x(\theta) = \bigg[x(0; \theta), \ldots, x\Big(\frac{kS}{m}; \theta\Big), \ldots, x\Big(\frac{(m{-}1)S}{m}; \theta\Big)\bigg]^T
\end{equation}
where $m$ is the number of sensors and $k \in \{0, 1, \ldots, m-1\}$. Additive
noise $\v \epsilon$ corrupts $\v x(\theta)$ to produce the noisy sensor values
\begin{equation} \label{eq:sensor-obs}
	\v y = \v x(\theta) + \v \epsilon.
\end{equation}
The noise $\v \epsilon$ is zero-mean and Gaussian, $\v \epsilon \sim
\mathcal{N}(\v 0, \m \Sigma)$. We restrict our analysis to a ``sensor noise''
setting, wherein $\m\Sigma = \sigma^2 \m I$, i.e., noise is i.i.d.\ across
sensors.

In the ``integrator sensor'' model, sensors have a \emph{width} which decreases
as their number grows to infinity. Concretely, the $k$-th sensor which is
located at $s = kS/m$ \emph{occupies} the space $((k{-}\frac{1}{2})S/m,
(k{+}\frac{1}{2})S/m)$.  The signal sensed by this sensor is given by the
$k$-th element of $\v x(\theta)$,
\begin{equation}
	x_k(\theta) = \int_{(k-\frac{1}{2})S/m}^{(k+\frac{1}{2})S/m} x(s;\theta) ds.
\end{equation}
In this model, the noise vector $\v \epsilon$ is also generated similarly, by
integrating over small sections of a white noise process, which is assumed to
additively corrupt the continuous-space signal $x(s;\theta)$ \emph{prior} to
sampling (see Fig.~\ref{fig:signal-model}). This leads to i.i.d.\ noise values
at sensors (since sensors have no overlap), with variance inversely
proportional to $m$, i.e., $\sigma(m) \propto 1/\sqrt{m}$.

In the source localization problem, we receive $n$ ``trials''\footnote{The word
	``trials'' comes from neuroscience literature, where multiple trials of an
	experiment are run, with the same stimulus being given to the participant
	in each trial, evoking the same response in the brain (with some
trial-to-trial variability, which we ignore). For our purposes, we assume that
trials are effectively i.i.d.\ realizations of sensor noise.} (i.e., $n$
realizations) of $\v y$ according to equation~\eqref{eq:sensor-obs}, for the
\emph{same} point source $\theta$, using which we need to estimate the location
of the source.  The noise is also assumed to be i.i.d.\ across trials, i.e.,
the $n$ realizations of $\v y$ are independently drawn from $P(\theta)$, for
the same location parameter $\theta$.

@@

Thus, the final lower bound can be written as
\begin{equation} \label{eq:le-cam-bound}
	\mathfrak{M}_n(\mathcal P, \Phiorho) \geq \frac{\Phi(\delta)}{2} \big(1 - \norm{P_1^n - P_0^n}_{TV}\big).
\end{equation}
The superscripts ``$n$'' remind us that these are $n$-fold product
distributions, since we have $n$ i.i.d.\ trials used in the estimate.  The
difficulty in Le Cam's method lies in selecting the two hypotheses to trade-off
the effect due to a small value of $\delta$ and a large value of $\norm{P_1^n -
P_0^n}_{TV}$ appropriately, to derive the tightest possible bound.

@@

@@

For the source localization problem, we measure \emph{risk} in terms of the
squared-distance between the true location and the estimated location. Hence,
our loss function is $\Phi(\rho(\theta, \hat\theta)) = \abs{\theta -
\hat\theta}^2$, where $\hat\theta$ is some estimate of the location based on
$n$ trials of the noisy sensor observations $\v y$. We now state the main
result of this paper.

@@

	\begin{equation*}

@@

	\end{equation*}

@@

section~\ref{sec:sensor-model}. Hence, combining

@@

	\norm{\v\Delta}^2 \leq 4 \kappa^2 \delta^2 \floor[\Big]{\!\frac{(w{+}2\delta)m}{S} + 1\!} \leq 4 \kappa^2 \delta^2 \bigg(\!\frac{(w{+}2\delta) m}{S} + 1\!\bigg).

@@

\subsection{Towards a more physical sensor model}
\label{sec:physical-model}

The bound in Theorem~\ref{thm:main-lb} may mislead one to think that the error
in source localization goes to zero, as the number of sensors goes to infinity.
This observation contradicts the information-theoretic lower bound given
in~\cite{Grover2016Fundamental}, where even with an infinite number of sensors,
the lower bound is strictly greater than zero. This indicates that the bound in
section~\ref{sec:lecam-lb} is failing to capture some effect which leads to
saturation of error as the number of sensors increases.

Indeed, the bound in Theorem~\ref{thm:main-lb} assumes that an arbitrarily
large number of sensors can be squeezed into a finite amount of space, and that
these sensors \emph{still} have the same SNR.  This is non-physical, since in
reality, each sensor has a certain width, and the value measured by it is the
integral of the continuous-space signal within that width (analogous to a
sample-and-hold circuit, which integrates a continuous-\emph{time} signal over
a short period). If the width of a sensor is decreased, then effectively, its
SNR reduces (this is explained in Appendix~\ref{app:integrator-snr} in the
full version of this paper~\cite{FullVersion}).

%This is intuitively evident in the sample-and-hold analogy: to compute the
%signal value, the integral must be divided by the time interval over which the
%integration is performed; this division amplifies noise as the time interval
%gets smaller.

We performed a preliminary study to understand the effect of using the
``integrator sensor'' model defined in section~\ref{sec:sensor-model}.  The
bounding technique used in section~\ref{sec:lecam-lb} no longer easily admits a
closed-form solution which gives a bound for all values of $m$. We instead
compute the bound numerically, using exact expressions for error probability in
a binary hypothesis test. For Gaussian distributions of equal variance and a
uniform prior on the hypotheses, the minimum possible error probability (over
all tests) is a $Q$-function in the distance between their
means~\cite[Ch.~7]{Scharf1991Statistical},
\begin{equation}
	P_e = \inf_\psi \mathbb P(\psi(Y^n) \neq V) = Q\bigg(\frac{\norm{\v\mu_0 - \v\mu_1}}{2\sigma(m)}\bigg).
\end{equation}
where $Q(x) = \int_x^\infty \frac{1}{\sqrt{2\pi}} e^{-u^2/2} du$. Hence, from
Theorem~\ref{thm:est-to-testing},
\begin{equation}
	\mathfrak{M}_n(\mathcal{P}, \Phiorho) \geq \Phi(\delta) P_e = \delta^2 Q\bigg(\frac{\norm{\v x(\theta_0) - \v x(\theta_1)}}{2\sigma(m)}\bigg).
\end{equation}
This bound is valid for any $\{\theta_0, \theta_1\}$ constituting a
$2\delta$-packing of $\Theta$. To numerically optimize over all possible
packings, we relax the process by fixing $\theta_0 = 0$ and optimizing over
$\theta_1 = 2\delta$.  We compute sensor values for a triangular impulse
response, by performing numerical integration, and find the best possible
$\delta$ by grid search.

\begin{figure}[t]
	\centering
	\includegraphics[width=2.8in]{lb-vs-m-annotated}
	\caption{A numerically generated plot of the lower bound on minimax risk of
		source localization, against the number of uniformly distributed
		sensors, $m$, used to estimate the location. The different curves are
		for different numbers of ``trials'', $n$ (number of i.i.d.\ sensor
		observations). The bound saturates for large $m$, indicating that for
		a \emph{fixed} number of trials, there is decreasing benefit to using
		more sensors.  However, with more trials, the bound saturates at
		increasingly larger numbers of sensors. Thus, there may be benefit to
	increasing the number of sensors when you have more trials.}
	\label{fig:numerical}
\end{figure}

Plotting this numerically computed bound for different numbers of trials (see
Fig.~\ref{fig:numerical}) reveals some interesting trends. First, the bound
saturates as $m$ grows large, matching the behaviour presented
in~\cite{Grover2016Fundamental}. Second, the lower bound on error is larger for
smaller numbers of sensors, as we expect. Third and most importantly, as the
number of trials increases, bound begins to saturate only for larger numbers of
sensors. In other words, there is increased benefit in using larger numbers of
sensors, when we also have more trials. This substantiates the intuition given
in~\cite{Grover2016Information}, which addresses this issue using a
Nyquist-rate perspective. A larger number of sensors will sample a larger
number of high-frequency components of the source signal (and hence allow us to
locate it better), but only if the noise floor is sufficiently low. Noise
variance, in turn, can be reduced by averaging over multiple trials.

@@

While the bounds in section~\ref{sec:lecam-lb} were derived for a
one-dimensional circular domain, extensions to a 1D linear domain (of fixed
length) and to a $d$-dimensional domain (circular or linear) are fairly
straightforward, under certain conditions. For a 1D linear domain, we require
that sensors capture the entirety of the impulse response even at the edges of
the domain. Hence, sensors must be placed uniformly in $[-w/2, S{+}w/2]$. For
extending to $d$ dimensions, it suffices to apply Assouad's
Lemma~\cite{Tsybakov2009Introduction} to separate the problem into $d$
independent 1D problems. Such extensions substantially broaden the number of
settings in which this bound applies.

Also note, the bounds derived in this paper apply only when sensors are
uniformly distributed.  Intuitively, this is the correct choice in the minimax
setting, since any non-uniform placement will have one or more points which is
far from all sensors. The ``maximization'' over $\theta \in \Theta$ would then
select a $\theta$ for which the value of the impulse response at the sensors is
smallest, so that SNR is minimized.

@@

\newpage

\appendices

\section{SNR in the integrator model}
\label{app:integrator-snr}
